---
title: "figure_1_comparisons"
author: "Julia Malnak"
date: "2025-06-09"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggExtra)
library(patchwork)
library(ggrastr)
```
# Figure 1 R-markdown

## Nucleotide distance

### loading in data
```{r}
mash_dist<-read_tsv("../intermediate_files/mash_distance_low_kmer.tsv", col_names=c("reference", "query", "distance", "p-value", "fraction"))
family_annotations<-read_tsv("../intermediate_files/family_species.tsv") %>% 
  drop_na() %>%
  distinct()
```

### adding in the metadata for mash distance so that I can change the labels
```{r}
mash_acc<- read_tsv("../intermediate_files/genomes_nomburg_represented.txt")
```

### Merge the two accessions
```{r}
mash_dist<- mash_dist %>% inner_join(mash_acc, by=c("reference"="Accession")) %>% rename(Reference_species=Species)
mash_dist <- mash_dist %>% inner_join(mash_acc, by=c("query"="Accession")) %>% rename(Query_species=Species)
```

```{r}
# filtering identities (self vs self) and if there is 
mash_dist <- mash_dist %>% select(Query_species, Reference_species, distance) %>% filter(Query_species!=Reference_species) %>%
  filter(!is.na(distance))
```

###Accounting for multiple reference genomes, I need to average each group
```{r}
mash_dist<- mash_dist  %>%
  rowwise() %>%
  mutate(pair = paste(sort(c(Reference_species, Query_species)), collapse = "_")) %>% # alphabetical
  ungroup() 
```

### Merge with the family and save for later analysis
```{r}
mash_fam <- mash_dist %>% 
  right_join(family_annotations, by=c("Query_species"="species")) %>%
  rename(query_family=family) %>%
  right_join(family_annotations, by=c("Reference_species"="species")) %>%
  rename(reference_family=family) %>%
  filter(query_family==reference_family) %>%
  filter(Query_species!=Reference_species) %>%
  distinct() %>%
  select(query_family, distance) %>%
    rename(family=query_family) %>%
    group_by(family) %>%
    summarise(diversity=mean(distance))
write_csv(mash_fam, "../results/family_diversity.csv")
```

### average the mash distance by pair
```{r}
avg_mash_dist<- mash_dist %>%
  group_by(pair)%>%
  mutate(avg_dist=mean(distance, na.rm=TRUE)) %>%
  ungroup() %>%
  select(-pair)
rm(mash_dist)
```


## Amino acid and nucleotide similarity

Loading in the structural similarity
Jaccard dissimilarity will be 1-jaccard- so if they are highly dissimilar they are at 1.
```{r}
## from mmseqs clustering
jaccard<-read_csv("../intermediate_files/species_jaccard.csv") %>% 
  select(-intersection, -union) %>% 
  mutate(jaccard_distance_structure=1-`jaccard index`)
## this is the old amino acid file
amino_acid<- read_csv("../intermediate_files/amino_acid_jaccard_distance.csv") %>% 
  select(-intersection, -union) %>%
  mutate(jaccard_distance_amino_acid=1-`jaccard index`) 

# joining the two df: amino acid and structure
jac_amin<- jaccard %>% inner_join(amino_acid, by=c("taxon1","taxon2")) %>%
  filter(taxon1!=taxon2)

```

### creating amino-acid vs nucleotide dataframe
```{r}
amino_mash <- amino_acid %>% 
  inner_join(avg_mash_dist, by=c("taxon1"="Query_species", "taxon2"="Reference_species"))
```

### creating the structure vs nuc
```{r}
jaccard_mash <- jaccard %>% 
  inner_join(avg_mash_dist, by=c("taxon1"="Query_species", "taxon2"="Reference_species"))
```
-
## Plot the distribution of the pairwise distances of the virus-virus relationships
```{r}
plot_dist<- function(df,dist_var,x)
{
  options(scipen=999)
  # save to a pdf 

df %>% 
  ggplot(mapping=aes(x={{dist_var}}))+
  geom_histogram(aes(y=..density..),fill="black")+
  theme_classic()+
    scale_x_reverse()+
    ylab("")+
    xlab("Pairwise Distance")+
  ylim(0, 30)

}
```

### Figure 1A, 1B. Supplementary Figure 1C
```{r}
# structure
plot_dist(jac_amin, jaccard_distance_structure,25)
# amino acid
plot_dist(jac_amin, jaccard_distance_amino_acid,25)
# nucleotide, takes the longest
plot_dist(avg_mash_dist,avg_dist, 30)

```

## Structure vs nucleotide
### Supplementary Figure 1D
### How similar is nucleotide similarity to structural similarity?
```{r}
# creating the dotplot structure 
p<-jaccard_mash %>% 
  ggplot(mapping=aes(x=avg_dist, y=jaccard_distance_structure))+
  rasterise(geom_point(alpha=0.25, size=1),dpi=300)+
  theme_classic()+
  xlab("Mash Distance")+
  ylab("Structure Distance")
p1<-ggMarginal(p, type="histogram", size=10, fill="black")
```

```{r}
# show plot
p1
```

## Structure vs amino acid
### Figure 1D
### How similar is amino acid similarity to structural similarity?
```{r}
#rasterization necessary for plot rendering
p<-jac_amin %>%
  ggplot(mapping =aes(x=jaccard_distance_amino_acid, y=jaccard_distance_structure))+
  rasterise(geom_point(alpha=0.05, size=1), dpi=300)+
  theme_classic()+
  xlab("Amino Acid Distance")+
  ylab("Structure Distance")
```

```{r}
p1<-ggMarginal(p, type="density", size=10, fill="black")
p1 # show the plot
```

## Amino acid vs nucleotide
## Supplementary 1A
### How is amino acid similarity compared to nucleotide distance?
```{r}
p<-amino_mash %>%
  ggplot(mapping=aes(x=avg_dist, y=jaccard_distance_amino_acid))+
  rasterise(geom_point(alpha=0.25, size=1), dpi=300)+
  theme_classic()+
  xlab("Mash Distance")+
  ylab("Amino Acid Distance")
p1<-ggMarginal(p, type="density", size=10, fill="black")
  
```

```{r}
# show plot
p1
```

## Intrafamilial distance
### Figure 1E
### Since it is the same metric, which families have greatest distance between amino acid similarity and structural similarity?
Within the family
```{r}
# remove redundant info from merging
jac_amin <- jac_amin %>% select(-`jaccard index.x`, -`jaccard index.y`)
```

```{r}
# merge the species with family labels
family_annotations<-read_tsv("../intermediate_files/family_species.tsv") %>% 
  drop_na() %>%
  distinct()
# add the family label to taxon 1
jac_amin<- jac_amin %>% inner_join(family_annotations, by=c("taxon1"="species")) %>%
  rename(family1=family)%>% inner_join(family_annotations, by=c("taxon2"="species")) %>%
  rename(family2=family)

```

```{r}
jac_amin<- jac_amin %>% 
  mutate(diff_str_am=jaccard_distance_structure-jaccard_distance_amino_acid)
```

```{r}
# pick the top35
top_families<- jac_amin %>% 
  filter(family1==family2) %>%
  distinct() %>%
  group_by(family1) %>%
  summarise(lg_diff=mean(diff_str_am))%>%
  arrange(lg_diff) %>%
  head(35) %>%
  pull(family1)
# plot the top 35
jac_amin %>% 
  filter(family1==family2) %>%
  distinct() %>%
  filter(family1 %in% top_families) %>%
  mutate(family1=factor(family1, levels=top_families, ordered = TRUE)) %>%
  #pivot_longer(cols=c("jaccard_distance_structure", "jaccard_distance_amino_acid"), names_to = "feature", values_to = "distance") %>%
  ggplot(mapping=aes(x=family1,y=diff_str_am))+
  rasterise(geom_jitter(alpha = 0.05, color="lightgrey",width=0.3), dpi=300)+
  geom_violin(fill=NA) +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  ylab("Jaccard Distance (Structure-Amino Acid)")+
  xlab("Viral Family")
 
```

#Supplementary Figure 1E
## Interfamilial similarity
Is there unexpected high similarity between families
```{r}
top_families<- jac_amin %>% 
  filter(family1!=family2) %>%
  distinct() %>%
  group_by(family1, family2) %>%
  summarise(lg_diff=mean(diff_str_am))%>%
  arrange(lg_diff) %>%
  head(60)%>%
  rowwise() %>%
  mutate(pair = paste(sort(c(family1, family2)), collapse = "-")) %>%
  distinct()%>%
  pull(pair) %>% unique()
jac_amin %>% 
  filter(family1!=family2) %>%
  rowwise() %>%
  mutate(pair = paste(sort(c(family1, family2)), collapse = "-")) %>%
  ungroup() %>%
  filter(pair %in% top_families) %>%
  select(pair,diff_str_am) %>%
  distinct() %>%
  mutate(pair=factor(pair, levels=top_families, ordered = TRUE)) %>%
  ggplot(mapping=aes(x=pair,y=diff_str_am))+
  rasterise(geom_jitter( color="grey",width=0.2), dpi=300)+
  geom_violin(fill=NA) +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  ylab("Jaccard Distance (Structure-Amino Acid)")+
  xlab("Viral Family Pairs")

  
```

Saving a dataframe of within family distance and between family for downstream analysis
```{r}
jac_amin %>% 
  filter(family1!=family2) %>%
  select(family1,family2,jaccard_distance_structure, diff_str_am) %>%
  write_csv("../results/between_family_dist.csv")
jac_amin %>% 
  filter(family1==family2) %>%
  select(family1,jaccard_distance_structure, diff_str_am) %>%
  write_csv("../results/within_family_dist.csv")
```



