---
title: "E4_analysis"
author: "Julia Malnak"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```

#Loading in Papillomavirus metadata
```{r}
E4_cluster_quality<- read_tsv("../intermediate_files/E4_analysis/E4_plddts.txt",
                              col_names = c("id", "plddt", "tropism")) 
E4_cluster_mems<- read_tsv("../intermediate_files/E4_analysis/cluster_mems.tsv",
                           col_names = c("id", "cluster_rep", "cluster_mem"))
```
## Looking at distribution of structure confidence with human and non-human information
```{r}
E4_cluster_quality %>% 
  ggplot(mapping = aes(x=plddt, fill=tropism))+
  geom_histogram(aes(y = after_stat(density)), color="#e9ecef", alpha=0.4, position = 'identity')+
  #geom_density(alpha=0.5)+
  scale_fill_manual(values=c("#404080", "#69b3a2" ))+
  theme_classic()
```
## Look at the distribution of the size of the clusters
# Supplementary Figure XXX
```{r}
E4_cluster_mems %>%
  left_join(E4_cluster_quality, by=c("cluster_rep"="id")) %>%
  select(-cluster_rep, -plddt) %>%
  group_by(id) %>%
  mutate(n=n()) %>%
  ungroup() %>%
  select(-cluster_mem) %>%
  distinct() %>%
  mutate(tropism=ifelse(id==699,"Human",tropism))%>%
  mutate(tropism=ifelse(id==5156,"No human",tropism))%>%
  ggplot(mapping=aes(x=n, fill=tropism))+
  geom_histogram()+
  xlab("Number of Members in Each E4 structural cluster")+
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10))+
  theme_classic()+
  scale_fill_manual(values=c("#404080", "#69b3a2" ))
```
#plddt of my clusters
```{r}
clust_id<-c(726,699,2522, 3232)
E4_cluster_mems %>% 
  filter(id %in% clust_id) %>% 
  left_join(E4_cluster_quality, by=c("cluster_mem"="id")) %>%
  distinct()

```
# Looking at the Sequence-based predictions of disorder in the N-terminal region

```{r}
# all sequence members with cluster ids
cluster_to_sequence<-read_tsv("../intermediate_files/E4_analysis/cluster_ids_all_members.tsv",
                              col_names=c("clust_id", "acc")) %>%
  group_by(clust_id) %>%
  mutate(mems=n()) %>%
  ungroup()
#IDR predictions
IUpred_long<-read_tsv("../intermediate_files/E4_analysis/E4_IUPred2A_long.tsv",
                      col_names=c("position", "AA", "IUPred2A_score", "ANCHOR2_score", "acc")) %>%
  select(-AA)
fldpnn<-read_tsv("../intermediate_files/E4_analysis/E4_fldpnn_results.csv", 
                 col_names = c(
                   "position","AA", "score","prediction",
                   "rdp_rna","rdp_dna","rdp_prot",
                   "dfl","molecular_recognition_feature", 
                   "acc"
                 ))
```
# Supplementary Figure XXX
```{r}
plot_scores<-function(df,score)
{
  # the dataframe is the scoring with acc and positon
  # score is the y-axis variable name
  score <- sym(score)
df %>%
  left_join(cluster_to_sequence, by="acc") %>%
  filter(mems >1) %>%
  ggplot(mapping=aes(x=position, y=!!score, group=acc))+
  geom_point(size=0.2)+
  geom_vline(xintercept = 20, color="red")+
  geom_line(linewidth=0.1)+
  facet_wrap(~clust_id)+
  theme_classic()
  
}
```
```{r}
plot_scores(IUpred_long,"IUPred2A_score")
plot_scores(IUpred_long, "ANCHOR2_score")
plot_scores(fldpnn, "score")
```

# Look at the distribution of scores for human vs non-human sequences
```{r}
# loading in human protein df
human_prots<- read_tsv("../intermediate_files/papilloma_human_prots.tsv", 
                       col_names = c("acc", "tropism"))

fldpnn_nterm<- fldpnn %>% 
  filter(position <=25) %>%
  group_by(acc) %>%
  summarise(mean_n_term_score=mean(score)) %>%
  left_join(cluster_to_sequence, by="acc") 

iupred_nterm<- IUpred_long %>% 
  filter(position <= 25) %>%
  group_by(acc) %>%
  summarise(mean_n_term_score=mean(IUPred2A_score))%>%
  left_join(cluster_to_sequence, by="acc") 
```

## Main Figure 4G
```{r}
#fldpnn
fldpnn_nterm %>%
  left_join(human_prots, by="acc") %>%
  ggplot(mapping=aes(x=tropism, y=mean_n_term_score, fill=tropism, color=tropism))+
  geom_boxplot(linewidth=1, alpha=0.1)+
  geom_jitter(size=2.5, fill="white")+
  theme_classic()+
  ylab("Mean fldpnn disorder score over first 25 amino acids")+
  scale_color_manual(values=c("#404080", "#69b3a2" ))+
  scale_fill_manual(values=c("#404080", "#69b3a2" ))
#IUPRED
iupred_nterm %>%
  left_join(human_prots, by="acc") %>%
  ggplot(mapping=aes(x=tropism, y=mean_n_term_score, fill=tropism, color=tropism))+
  geom_boxplot(linewidth=1, alpha=0.1)+
  geom_jitter(size=2.5, fill="white")+
  theme_classic()+
  ylab("Mean IUPred disorder score over first 25 amino acids")+
  scale_color_manual(values=c("#404080", "#69b3a2" ))+
  scale_fill_manual(values=c("#404080", "#69b3a2" ))
  
```

```{r}

# fldpnn
hu <- fldpnn_nterm  %>% 
  left_join(human_prots, by="acc") %>%
  filter(tropism=="Human") %>%
  pull(mean_n_term_score)
# pull non-human
non_hu <- fldpnn_nterm  %>% 
  left_join(human_prots, by="acc") %>%
  filter(tropism!="Human") %>%
  pull(mean_n_term_score)
wilcox.test(hu,non_hu)
# IUPRED
# REPEAT for fldpnn
hu <- iupred_nterm  %>% 
  left_join(human_prots, by="acc") %>%
  filter(tropism=="Human") %>%
  pull(mean_n_term_score)
# pull non-human
non_hu <- iupred_nterm  %>% 
  left_join(human_prots, by="acc") %>%
  filter(tropism!="Human") %>%
  pull(mean_n_term_score)
wilcox.test(hu,non_hu)

```

