---
title: "human_visual_specific"
author: "Julia Malnak"
date: "2025-09-22"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Code for visualizing figure 4c
The figure comes from the human_spec.py file and is the results saved from
positve-negative set, where the sets are defined as human prevalent or prevalent in non-human.
- means present in non-human and + means present in human

### Loading in the libraries and data
```{r}
library(tidyverse)
library(paletteer)

df <- read_csv("../intermediate_files/prevalence_change_analysis.csv") %>% 
  rename(family= "...1")
# filter out circo, genomo, and smacoviridae because the effect is so small
filter_virus<-c("Circoviridae", "Genomoviridae", "Smacoviridae")
# filter if the row is 0
df <- df %>% 
  rowwise() %>%
  mutate(sum_prev=sum(c_across(-family), na.rm = TRUE)) %>%
  filter(sum_prev !=0) %>%
  select(-sum_prev) %>%
  filter(! family %in% filter_virus)

# loading in the primate specific results 
primate_df<-read_csv("../intermediate_files/primate_prevalence_change_analysis.csv")%>% 
  rename(family= "...1")
primate_df <- primate_df%>% 
  rowwise() %>%
  mutate(sum_prev=sum(c_across(-family), na.rm = TRUE)) %>%
  filter(sum_prev !=0) %>%
  select(-sum_prev) %>%
  filter(! family %in% filter_virus)
```

### Show the data as a heatmap
```{r}
library(pheatmap)
plot_heatmap<-function(data)
{

mat <- data %>%
  select(-"27", -"2170") %>% # these were not strongly enriched features
  column_to_rownames("family") %>%
  as.matrix()
# setting the colors
col <- colorRampPalette(c("#075AFFFF", "white","#FF0000FF"))(29)
breaks_list <- seq(-1, 1, length.out = 31)
pheatmap(mat,
         color = col,
         breaks=breaks_list,
         cellwidth = 30,
         cellheight = 40)
}
```

```{r}
plot_heatmap(df)
plot_heatmap(primate_df)
```

#Examine the differences
```{r}
# human specific
human_mat <- df %>% select(colnames(primate_df)) %>%
  mutate(family= factor(family, levels = primate_df$family)) %>%
  arrange(family) %>%
  column_to_rownames("family") %>%
  as.matrix()
primate_mat <- primate_df %>%
  mutate(family= factor(family, levels = primate_df$family)) %>%
  arrange(family) %>%
  column_to_rownames("family") %>%
  as.matrix()
diff<- primate_mat-human_mat

# plot
col <- colorRampPalette(c("#075AFFFF", "white","#FF0000FF"))(29)
breaks_list <- seq(-0.1, 0.1, length.out = 31)
pheatmap(diff,
         color = col,
         breaks=breaks_list,
         cellwidth = 30,
         cellheight = 40,
         cluster_cols = FALSE,
         show_colnames = TRUE,
         show_rownames=TRUE)
  
```



### Examine where 32/104 is not present in human vs non-human infecting viruses
```{r}
# loading the 32, 104 filtered df
feature_vec<- read_csv("../intermediate_files/32_154_clust_frequency.csv")
# loading in the species to family mapping
fam_species<-read_tsv("../intermediate_files/family_species.tsv")
# merge and only keep Poxviridae, Adenoviridae, Herpesviridae
feature_vec<- feature_vec %>%
  inner_join(fam_species, by="species") %>%
  filter(family=="Adenoviridae" | 
           family=="Poxviridae" |
           family=="Herpesviridae") %>%
  distinct()
```

Plot which virus families are missing some of the 32 or 104 features
```{r}
## feature vec is master table do not touch
vals<- feature_vec %>%
  select(species, `32`, `154`) %>%
  as.data.frame()

# add annotations
row_anno<- feature_vec %>% 
  select(family,Tropism, species) %>%
  column_to_rownames("species") %>%
  as.data.frame()
# add annotation colors
row_colors=list(
  family=c(Herpesviridae="#A1D8B6FF", 
           Adenoviridae="#D2C48EFF",
           Poxviridae="#F45F40FF"),
  Tropism=c(`Non-human`="#F9AE8DFF",
    Human="#80B9CEFF")
  )

# transform the df into a matrix
mat_vals <- vals %>%
  column_to_rownames("species") %>%
  as.matrix()
pheatmap(mat_vals,
         color = colorRampPalette(c("white", "black"))(50),
         border_color="black",
         annotation_row = row_anno,
         annotation_colors=row_colors,
         cluster_cols = FALSE, 
         cellwidth = 10,
         cellheight = 10)

```

```{r}
# look at the proportion of clusters in human vs non-human infecting clusters
feature_vec %>%
  pivot_longer(cols=c(`32`,`154`),
               names_to = "cluster_ID",
               values_to= "presence") %>%
  group_by(family,Tropism,cluster_ID) %>%
  mutate(mems=n()) %>%
  mutate(clust_presences=sum(presence)) %>%
  ungroup() %>%
  mutate(prop=clust_presences/mems) %>%
  select(family,prop,cluster_ID,Tropism, clust_presences,mems) %>%
  distinct() %>%
  ggplot(mapping=aes(x=family,y=prop, group=Tropism, fill=Tropism))+
  geom_bar(position=position_dodge(),
           stat="identity")+
  facet_wrap(~cluster_ID)+
  theme_classic()+
  scale_fill_manual(values=c("#404080", "#69b3a2" ))
```

### now test for significant association in cluster 32
```{r}
contigen<- feature_vec %>% 
  distinct() %>%
  mutate(clust_present=ifelse(`32`==1, "Present", "Absent")) %>%
  select(clust_present,Tropism) 
contigen <- table(contigen$clust_present, contigen$Tropism)
chisq.test(contigen)
```

## Looking for differences in clusters 32 and 665 
###Length
```{r}
length_32_clust<- read_tsv("../intermediate_files/length_clust_32_prot.tsv",
                           col_names = c("Acc", "length"))
length_32_clust %>%
  ggplot(mapping=aes(x=length))+
  geom_histogram(color="white",fill="black", bins=20)+
  geom_vline(xintercept=248, color="red")+
  geom_vline(xintercept = 262, color="red")+
  theme_classic()
```

###plddt
```{r}
plddt_32_clust<- read_tsv("../intermediate_files/plddts_cluster_32.tsv",
                          col_names = c("clust_rep", "plddt"))
plddt_32_clust %>%
  ggplot(mapping=aes(x=plddt))+
  geom_histogram(color="white", fill="black", bins=20)+
  geom_vline(xintercept =	86.21, color="red")+
  geom_vline(xintercept= 83.85, color="red")+
  theme_classic()
```

#Examining pfam related clusters
```{r}
# filter for the families where these clusters are present
present_family<-c("Herpesviridae", "Poxviridae", "Adenoviridae", "Alloherpesviridae")
fam_species<- fam_species %>% filter(family %in% present_family)
# load in the data and filter relevant families only
pfam_clust_32<- read_csv("../intermediate_files/clust_32_analysis/frequent_clust_32_pfam.csv") %>%
  inner_join(fam_species, by="species") %>%
  distinct()

```

# first look at the prportion of viruses that contain any igfold
```{r}
pfam_clust_any_ig<- pfam_clust_32 %>%
  pivot_longer(cols=`1051`:`8944`,
               names_to = "cluster_id",
               values_to = "presence") %>%
  group_by(species) %>%
  mutate(present_once=ifelse(sum(presence)>0, 1, 0)) %>%
  ungroup() %>%
  select(species, family,Tropism, present_once) %>%
  group_by(family,Tropism) %>%
  distinct() %>%
  summarise(mems=n(),
         present_trop=sum(present_once),
         prop=present_trop/mems) 
# plot, family specific
pfam_clust_any_ig %>%
ggplot(mapping=aes(x=family,y=prop, group=Tropism, fill=Tropism))+
  geom_bar(position=position_dodge(),
           stat="identity")+
  theme_classic()+
  scale_fill_manual(values=c("#404080", "#69b3a2" ))
# plot all together
pfam_clust_any_ig %>%
  ungroup() %>%
  select(Tropism, mems, present_trop) %>%
  group_by(Tropism) %>%
  summarise(infecting=sum(mems),
         ig_present=sum(present_trop),
         prop=ig_present/infecting) %>%
ggplot(mapping=aes(x=Tropism,y=prop, fill=Tropism))+
  geom_bar(position=position_dodge(),
           stat="identity")+
  theme_classic()+
  scale_fill_manual(values=c("#404080", "#69b3a2" ))
```

Look at if there is a cluster specific driver
```{r}
family_specific_cluster<-pfam_clust_32 %>%
  pivot_longer(cols=`1051`:`8944`,
               names_to = "cluster_ID",
               values_to = "present") %>%
  distinct() %>%
  group_by(family,Tropism,cluster_ID) %>%
  summarise(mems=n(),
    present_in_clust=sum(present),
         prop=present_in_clust/mems)
# plot the results
family_specific_cluster %>%
  ggplot(mapping=aes(x=family,y=prop, group=Tropism, fill=Tropism))+
  geom_bar(position=position_dodge(),
           color="white",
           stat="identity")+
  theme_classic()+
  scale_fill_manual(values=c("#404080", "#69b3a2" ))+
  facet_wrap(~cluster_ID)+
  theme(axis.text.x =element_text(angle=90))

  
```

## Proportion of all human-infecting vs non human infecting viruses than contain a sequence cluster of this fold
```{r}
ig_containing_viruses<-read_tsv("../intermediate_files/clust_32_analysis/viruses_containing_seq_cluster_with_ig_domain.tsv",
                                col_names=c("species","taxid")) %>%
  mutate(contain_ig=TRUE)
all_viruses<-read_tsv("../intermediate_files/clust_32_analysis/no_ig_domain.tsv", 
                      skip=2,
                      col_names = c("species", "taxid")) %>%
  mutate(contain_ig=FALSE)
viruses<- all_viruses %>% rbind(ig_containing_viruses) %>% distinct()%>%
  filter(!is.na(species))
# loading the human annotations
hu<- read_tsv("../intermediate_files/human_included_viruses.csv" , 
              col_names=c("species")) %>%
  pull(species)

# add Vaccinia, not included
hu<-c(hu, "Vaccinia_virus")

viruses <- viruses %>% mutate(Tropism=ifelse(species %in% hu, "Human", "Non-human"))
# proportion of human infecting vs non-human infecting containing a ig-domain

viruses_summary <- viruses %>%
  group_by(Tropism) %>%
  summarise(mems=n(),
            ig_containing=sum(contain_ig),
            prop=ig_containing/mems)

# plot the results
viruses_summary %>%
  ggplot(mapping=aes(x=Tropism,
                     y=prop, fill=Tropism))+
  geom_bar(position=position_dodge(),
           color="white",
           stat="identity")+
   theme_classic()+
  scale_fill_manual(values=c("#404080", "#69b3a2" ))
```

# Look at this in structural clusters
```{r}
ig_containing_virus_str<-read_tsv("../intermediate_files/clust_32_analysis/viruses_with_structure_ig.tsv",
                                  col_names=c("species","taxid")) %>%
  mutate(contain_ig=TRUE)
no_ig_containing_virus_str<- read_tsv("../intermediate_files/clust_32_analysis/viruses_no_ig_structural_cluster.tsv",
                                      skip=2,
                                      col_names=c("species","taxid")) %>%
  mutate(contain_ig=FALSE)
# stack into one large df
viruses_str<- ig_containing_virus_str %>% rbind(no_ig_containing_virus_str) %>% 
  distinct()%>%
  filter(!is.na(species))
# adding in infectivity information
viruses_str <- viruses_str %>% mutate(Tropism=ifelse(species %in% hu, "Human", "Non-human"))
virus_summary_str<-  viruses_str %>%
  group_by(Tropism) %>%
  summarise(mems=n(),
            ig_containing=sum(contain_ig),
            prop=ig_containing/mems)
virus_summary_str %>%
  ggplot(mapping=aes(x=Tropism,
                     y=prop, fill=Tropism))+
  geom_bar(position=position_dodge(),
           color="white",
           stat="identity")+
   theme_classic()+
  scale_fill_manual(values=c("#404080", "#69b3a2" ))
```

# continue this analysis with looking at the family distribution of pfam terms
```{r}
ig_containing_clust_info<- read_tsv("../intermediate_files/clust_32_analysis/ig_containing_cluster_lines.tsv",
                                    col_names =c("clust_id", "str_clust_rep", "seq_clust_rep", "subclust_mem", "?", "Realm", "Phylum",
                                                 "Class", "Order", "Family", "Genus", "species", "taxid") )
min_ig_containing<- ig_containing_clust_info %>% select(clust_id,species) %>%
  mutate(ig_containing=TRUE)
filter_fam<-unique(ig_containing_clust_info$Family)
fam_species<-read_tsv("../intermediate_files/family_species.tsv") %>%
  filter(family %in% filter_fam) 
min_ig_containing<- fam_species %>%
  left_join(min_ig_containing, by="species") %>%
  distinct()  %>% 
  mutate(Tropism=ifelse(species %in% hu, "Human", "Non-human"))
```
# look at family level distribution (how many in each family contain the cluster?)
```{r}
fam_ig_contain <- min_ig_containing %>%
  mutate(ig_containing=replace_na(ig_containing,FALSE)) %>%
  group_by(species) %>%
  mutate(present=ifelse(sum(ig_containing) >0, 1, 0)) %>%
  ungroup() %>%
  select(family, species, Tropism, present) %>%
  distinct() %>%
  group_by(family,Tropism) %>%
  summarise(mems=n(),
            ig_present=sum(present),
            prop=ig_present/mems)
fam_ig_contain %>%
  ggplot(mapping=aes(x=family,y=prop, group=Tropism, fill=Tropism))+
  geom_bar(position=position_dodge(),
           color="white",
           stat="identity")+
  theme_classic()+
  scale_fill_manual(values=c("#404080", "#69b3a2" ))+
  theme(axis.text.x =element_text(angle=90))
            
```

# look at the cluster distribution
```{r}
cluster_ig_containing<- min_ig_containing %>%
  distinct() %>%
  group_by(family,Tropism) %>%
  mutate(species_count=n_distinct(species)) %>%
  group_by(family,Tropism,clust_id) %>%
  mutate(clust_present=sum(ig_containing),
         prop=clust_present/species_count) 

cluster_ig_containing %>%
  filter(!is.na(clust_id)) %>%
  ggplot(mapping=aes(x=family,y=prop, group=Tropism, fill=Tropism))+
  geom_bar(position=position_dodge(),
           color="white",
           stat="identity")+
  theme_classic()+
  scale_fill_manual(values=c("#404080", "#69b3a2" ))+
  facet_wrap(~clust_id)+
  theme(axis.text.x =element_text(angle=90)) 
```

# look at the length of the proteins in the cluster
```{r}
rep_clust_id <- ig_containing_clust_info %>% select(clust_id, seq_clust_rep)
prot_length<-read_tsv("../intermediate_files/clust_32_analysis/protein_length_ig_containing_str.tsv", 
                      col_names = c("acc", "len"))
prot_clust<-read_tsv("../intermediate_files/clust_32_analysis/clust_acc_id.tsv",
                     col_names = c("id", "acc"))
 prot_length %>% 
  left_join(prot_clust, by="acc")  %>%
  mutate(id=as.factor(id)) %>%
  group_by(id) %>%
  summarise(mean=mean(len),
            sd=sd(len)) %>%
  ggplot(mapping = aes(x=id, y=mean))+
  geom_col()+
  geom_errorbar(aes(ymin=pmax(0,mean-sd),
                    ymax=mean+sd),
                width=0.2)+
  theme_classic()+
  ylab("Length")+
  xlab("Cluster ID")+
  theme(axis.text.x =element_text(angle=90)) 
```

# look at the plddt of the clusters
```{r}
rep_clust_id <- ig_containing_clust_info %>% select(clust_id, seq_clust_rep)
prot_plddt<-read_tsv("../intermediate_files/clust_32_analysis/plddts_ig_containing_structures.tsv", col_names = 
                       c("seq_clust_rep", "plddt"))  %>%
  left_join(rep_clust_id, by="seq_clust_rep") %>% distinct()
prot_plddt %>%
  mutate(clust_id=as.factor(clust_id)) %>%
  group_by(clust_id) %>%
  summarise(mean=mean(plddt),
            sd=sd(plddt)) %>%
  ggplot(mapping = aes(x=clust_id, y=mean))+
  geom_col()+
  geom_errorbar(aes(ymin=mean-sd,
                    ymax=mean+sd),
                width=0.2)+
  theme_classic()+
  ylab("Plddt")+
  xlab("Cluster ID")+
  theme(axis.text.x =element_text(angle=90)) 
```


