---
title: "human_visual_specific"
author: "Julia Malnak"
date: "2025-09-22"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Code for visualizing figure 4c
The figure comes from the human_spec.py file and is the results saved from
positve-negative set, where the sets are defined as human prevalent or prevalent in non-human.
- means present in non-human and + means present in human

### Loading in the libraries and data
```{r}
library(tidyverse)
library(paletteer)

df <- read_csv("../results/prevalence_change_analysis.csv") %>% 
  rename(family= "...1")
# filter out circo, genomo, and smacoviridae because the effect is so small
filter_virus<-c("Circoviridae", "Genomoviridae", "Smacoviridae")
# filter if the row is 0
df <- df %>% 
  rowwise() %>%
  mutate(sum_prev=sum(c_across(-family), na.rm = TRUE)) %>%
  filter(sum_prev !=0) %>%
  select(-sum_prev) %>%
  filter(! family %in% filter_virus)
  

```
### Show the data as a heatmap
```{r}
library(pheatmap)
mat <- df %>%
  select(-"27", -"2170") %>% # these were not strongly enriched features
  column_to_rownames("family") %>%
  as.matrix()
# setting the colors
col <- colorRampPalette(c("#075AFFFF", "white","#FF0000FF"))(29)
breaks_list <- seq(-1, 1, length.out = 31)
pheatmap(mat,
         color = col,
         breaks=breaks_list,
         cellwidth = 30,
         cellheight = 40)
```

### Examine where 32/104 is not present in human vs non-human infecting viruses
```{r}
# loading the 32, 104 filtered df
feature_vec<- read_csv("../intermediate_files/32_154_clust_frequency.csv")
# loading in the species to family mapping
fam_species<-read_tsv("../intermediate_files/family_species.tsv")
# merge and only keep Poxviridae, Adenoviridae, Herpesviridae
feature_vec<- feature_vec %>%
  inner_join(fam_species, by="species") %>%
  filter(family=="Adenoviridae" | 
           family=="Poxviridae" |
           family=="Herpesviridae") %>%
  distinct()
```

Plot which virus families are missing some of the 32 or 104 features
```{r}
## feature vec is master table do not touch
vals<- feature_vec %>%
  select(species, `32`, `154`) %>%
  as.data.frame()

# add annotations
row_anno<- feature_vec %>% 
  select(family,Tropism, species) %>%
  column_to_rownames("species") %>%
  as.data.frame()
# add annotation colors
row_colors=list(
  family=c(Herpesviridae="#A1D8B6FF", 
           Adenoviridae="#D2C48EFF",
           Poxviridae="#F45F40FF"),
  Tropism=c(`Non-human`="#F9AE8DFF",
    Human="#80B9CEFF")
  )

# transform the df into a matrix
mat_vals <- vals %>%
  column_to_rownames("species") %>%
  as.matrix()
pheatmap(mat_vals,
         color = colorRampPalette(c("white", "black"))(50),
         border_color="black",
         annotation_row = row_anno,
         annotation_colors=row_colors,
         cluster_cols = FALSE, 
         cellwidth = 10,
         cellheight = 10)

```

```{r}
# look at the proportion of clusters in human vs non-human infecting clusters
feature_vec %>%
  pivot_longer(cols=c(`32`,`154`),
               names_to = "cluster_ID",
               values_to= "presence") %>%
  group_by(family,Tropism,cluster_ID) %>%
  mutate(mems=n()) %>%
  mutate(clust_presences=sum(presence)) %>%
  ungroup() %>%
  mutate(prop=clust_presences/mems) %>%
  select(family,prop,cluster_ID,Tropism, clust_presences,mems) %>%
  distinct()
  ggplot(mapping=aes(x=family, fill=presence, group=Tropism))+
  geom_bar(position=position_dodge(),
           stat=after_stat(prop))+
  facet_wrap(~cluster_ID)+
  theme_classic()
```


