---
title: "KNN_analysis_vs_amino_acids"
author: "Julia Malnak"
date: "2025-06-05"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(paletteer)
library(ggrastr)
library(patchwork)
library(ComplexHeatmap)
library(circlize)
library(RColorBrewer)
```

## Comparing KNN performance across Structure-based Cluster and Amino-Acid Clusters
```{r}
am_knn_3<-read_csv("../intermediate_files/results_knn/knn_result_k_3_amino_acids.csv") 
am_knn_5<-read_csv("../intermediate_files/results_knn/knn_result_k_5_amino_acids.csv") 
am_knn_11<-read_csv("../intermediate_files/results_knn/knn_result_k_11_amino_acids.csv")
str_knn_3<-read_csv("../intermediate_files/results_knn/knn_result_k_3_structure.csv")%>% mutate(k=3) %>% mutate(feature="structure")
str_knn_5<- read_csv("../intermediate_files/results_knn/knn_result_k_5_structure.csv")%>% mutate(k=5) %>% mutate(feature="structure")
str_knn_11<-read_csv("../intermediate_files/results_knn/knn_result_k_11_structure.csv")%>% mutate(k=11) %>% mutate(feature="structure")
```

```{r}
filter_species<- function(knn_df, shared_species)
{
  # in a function in case the code needs to change
  knn_df<- knn_df %>% filter(species %in% shared_species)
  return(knn_df)
}
```

```{r}
# make sure they only compared the same set of things
knn_3_species<- intersect(am_knn_3$species, str_knn_3$species)
knn_5_species<- intersect(am_knn_5$species, str_knn_5$species)
knn_11_species<-intersect(am_knn_11$species, str_knn_11$species)

# filter out the species that don't match
am_knn_3<- filter_species(am_knn_3, knn_3_species)  
am_knn_5<- filter_species(am_knn_5, knn_5_species)
am_knn_11<- filter_species(am_knn_11, knn_11_species)
str_knn_3<- filter_species(str_knn_3, knn_3_species)
str_knn_5<- filter_species(str_knn_5, knn_5_species)
str_knn_11<- filter_species(str_knn_11, knn_11_species)
```

```{r}
# need to merge to make sure the species tested are the same for both sets
accuracy<-function(knn_df)
{
  knn_acc<- knn_df %>% group_by(known_family) %>%
  summarise(correct=sum(known_family==predicted_family),
         overall=n()) %>%
  mutate(accuracy=correct/overall) %>%
  mutate(known_family = reorder(known_family, -accuracy))
  
  return(knn_acc)
}

```

```{r}
# add amino acids
am_knn_3<-accuracy(am_knn_3)%>% mutate(k=3) %>% mutate(feature="Amino Acid") %>% mutate(run="Amino Acid 3")
am_knn_5<-accuracy(am_knn_5)%>% mutate(k=5) %>% mutate(feature="Amino Acid") %>% mutate(run="Amino Acid 5")
am_knn_11<-accuracy(am_knn_11)%>% mutate(k=11) %>% mutate(feature="Amino Acid")%>% mutate(run="Amino Acid 11")
# add structure
str_knn_3<-accuracy(str_knn_3) %>% mutate(k=3) %>% mutate(feature="Structure") %>% mutate(run="Structure 3")
str_knn_5<-accuracy(str_knn_5) %>% mutate(k=5) %>% mutate(feature="Structure") %>% mutate(run="Structure 5")
str_knn_11<-accuracy(str_knn_11) %>% mutate(k=11) %>% mutate(feature="Structure")%>% mutate(run="Structure 11")
```

```{r}
# merge all together
knn <- am_knn_3 %>% 
  bind_rows(am_knn_5) %>% 
  bind_rows(am_knn_11) %>% 
  bind_rows(str_knn_3) %>%
  bind_rows(str_knn_5) %>%
  bind_rows(str_knn_11) 
```

```{r}
knn<- knn %>% mutate(k=as.factor(k))
palette <- paletteer::paletteer_d("rcartocolor::TealRose")
sub_color<-c(palette[1], palette[length(palette)])
```
## Supplementary Figure 3B
Let's look at the distribution of accuracy
```{r}
ggplot(knn, aes(x = k, y = accuracy, color = feature)) +
  geom_jitter(alpha = 0.5, position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.9)) +
  geom_boxplot(fill=NA, position = position_dodge(width = 0.9), outliers=FALSE) +
  theme_classic()+
  scale_color_manual(values=sub_color, name="Cluster Type")+
  ylab("Family Accuracy")+
  xlab("K")

```



## Figure correlating accuracy to the discrepancy between the structural and amino acid distances
```{r}
#loading in the dataframe of interest
within_family<-read_csv("../intermediate_files/family_amino_acid_structure_distance/within_family_dist.csv") %>%
  mutate(comparision="within family") 
between_family<-read_csv("../intermediate_files/family_amino_acid_structure_distance/between_family_dist.csv") %>%
  mutate(comparision="between family")
diversity<-read_csv("../intermediate_files/family_amino_acid_structure_distance/family_diversity.csv")
```

```{r}
knn_no_na<- knn %>% select(-feature, -k, -correct, -overall) %>% distinct() %>% pivot_wider(names_from = "run", values_from = "accuracy") %>%
  drop_na()
knn_subset<- knn_no_na %>% select(known_family,`Structure 11`)
between_family<- between_family %>% inner_join(knn_subset, by=c("family1"="known_family"))
within_family<- within_family %>% 
  inner_join(knn_subset, by=c("family1"="known_family")) %>% 
  mutate(family2="None") 
family_comparision<- between_family %>% 
  rbind(within_family) %>% 
  rename(Accuracy=`Structure 11`) %>%
  mutate(jaccard_similarity=1-jaccard_distance_structure) %>%
  select(-jaccard_distance_structure)
within_family <- within_family %>%
  inner_join(diversity, by=c("family1"="family"))
```


# Does the difference in accuracy between amino acid and structure correlate with performance?
```{r}
am_mini_11<- am_knn_11 %>% select(known_family, accuracy) %>% rename(amino_acid_accuracy=accuracy)
str_mini_11<- str_knn_11 %>% select(known_family, accuracy) %>% rename(structure_accuracy=accuracy)

merged_diff <- str_mini_11 %>% inner_join(am_mini_11, by="known_family") %>%
  inner_join(diversity, by=c("known_family"="family")) %>%
  mutate(diff_accuracy=amino_acid_accuracy-structure_accuracy) 

merged_diff %>% 
  ggplot(mapping=aes(x=diff_accuracy, y=diversity))+
  geom_point()+
  theme_classic()+
  xlab("Amino Acid Accuracy- Structural Accuracy")+
  ylab("Diversity")
```

## Figure 2E
```{r}
# prep dataframe for visualization
family_comparision<- family_comparision %>%
  inner_join(am_mini_11, by=c("family1"="known_family")) %>%
  mutate(diff_acc=amino_acid_accuracy-Accuracy) %>%
    mutate(family1 = as.factor(family1),                     
         family1 = fct_reorder(family1, diff_acc))
```

 Let's stratify this figure into a geom tile and distribution figure
```{r}
tile<-family_comparision %>% 
  select(family1,diff_acc) %>%
    mutate(acc="Sensitivity") %>%
  distinct() %>%
  ggplot(mapping=aes(y=family1, x=acc, fill=diff_acc))+
  geom_tile(linewidth=0.5, color="black")+
  scale_fill_gradient2(low="#075AFFFF",
                          mid="white",
                          high="#FF6600FF",
                          midpoint=0)+
  theme_classic()+
  coord_fixed(ratio=0.5)+
  theme(legend.position = "left")
  

box_p<-family_comparision %>%
  ggplot(mapping = aes(x=diff_str_am, y=family1))+
  geom_boxplot(
     aes(group = interaction(family1, comparision), fill=comparision),
    position = position_dodge(width = 0.9),
    outlier.shape = NA, alpha=0.7, linewidth=0.15)+
  scale_fill_manual(values=c("#046E8FFF","#D44D5CFF"))+
  theme_classic()+
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(), 
    axis.ticks.y=element_blank()
  )+
  xlab("Jaccard Distance (Structure-Amino Acid)")

# put together using patchwork
tile +box_p +plot_layout(ncol=2)
```



## Looking at all taxonomic levels
# Benchmarking all taxonomic levels!
```{r}
# adding in the knn_results<- 
genus<-read_csv("../intermediate_files/results_knn/knn_result_k_11_structure_genus.csv") 
family<-read_csv("../intermediate_files/results_knn/knn_result_k_11_structure.csv") %>% 
  rename(known_taxa=known_family) %>% 
  rename(predicted_taxa=predicted_family)
order<-read_csv("../intermediate_files/results_knn/knn_result_k_11_structure_order.csv")
class<-read_csv("../intermediate_files/results_knn/knn_result_k_11_structure_class.csv")
phyla<-read_csv("../intermediate_files/results_knn/knn_result_k_11_structure_phyla.csv")

```

```{r}
mod_accuracy<-function(knn_df)
{
  knn_acc<- knn_df %>% group_by(known_taxa) %>%
  mutate(correct=sum(known_taxa==predicted_taxa),
         overall=n()) %>%
  mutate(accuracy=correct/overall) %>%
  mutate(known_taxa = reorder(known_taxa, -accuracy)) %>%
    ungroup() %>%
    select(species,accuracy)
  
  return(knn_acc)
}
```


```{r}
genus<-mod_accuracy(genus)
family<-mod_accuracy(family)
order<-mod_accuracy(order)
class<-mod_accuracy(class)
phyla<-mod_accuracy(phyla)
```

```{r}
taxa_guide <- read_tsv("../intermediate_files/taxonomic_levels.tsv") 
# merge each species accuarcy and rename with the taxonomic level
taxa_guide <- taxa_guide %>% left_join(phyla, by="species") %>% rename(phylum_accuracy=accuracy)
taxa_guide <- taxa_guide %>% left_join(family, by="species") %>% rename(family_accuracy=accuracy)
taxa_guide<- taxa_guide %>% left_join(order, by="species") %>% rename(order_accuracy=accuracy)
taxa_guide<- taxa_guide %>% left_join(class, by="species") %>% rename(class_accuracy=accuracy)
taxa_guide<- taxa_guide %>% left_join(genus, by="species") %>% rename(genus_accuracy=accuracy) %>% distinct()
# tier by phyla, class, order, family, genus
taxa_guide<- taxa_guide %>% 
  filter(!is.na(species)) %>% 
  filter(!if_all(c(phylum_accuracy, class_accuracy, order_accuracy,family_accuracy, genus_accuracy), is.na)) %>%
  arrange(phylum, class, order, family, genus) 
# look at the accuracy across all levels
taxa_guide %>% 
  summarise(across(everything(), ~ mean(.x, na.rm = TRUE)))
# now group by phyla and average the accuracy across all levels
# phyla accuracy
P<-taxa_guide %>% 
  group_by(phylum) %>%
  summarise(mean(phylum_accuracy, na.rm=TRUE))
# order accuracy
O<-taxa_guide %>%
  select(phylum, order, order_accuracy) %>%
  distinct()%>%
  ungroup() %>%
  group_by(phylum) %>%
  summarise(mean(order_accuracy, na.rm=TRUE))
# class accuracy
C<-taxa_guide %>%
  select(phylum, class, class_accuracy) %>%
  distinct()%>%
  ungroup() %>%
  group_by(phylum) %>%
  summarise(mean(class_accuracy, na.rm=TRUE))
# family accuracy
FAM<-taxa_guide %>%
  select(phylum, family, family_accuracy) %>%
  distinct()%>%
  ungroup() %>%
  group_by(phylum) %>%
  summarise(mean(family_accuracy, na.rm=TRUE))
# genus accuracy
g<-taxa_guide %>%
  select(phylum, genus, genus_accuracy) %>%
  distinct()%>%
  ungroup() %>%
  group_by(phylum) %>%
  summarise(mean(genus_accuracy, na.rm=TRUE))
## merge into one table
g %>% 
  left_join(FAM, by="phylum") %>%
  left_join(C, by="phylum") %>% 
  left_join(O, by="phylum") %>%
  left_join(P, by="phylum")
# look at the minimum accuracy
taxa_guide %>% 
  group_by(phylum) %>%
  select(-class, -order, -family, -genus, -species) %>%
  summarise(across(everything(), ~ min(.x, na.rm = TRUE)))
# look at average performance in each taxonomic level
taxa_guide %>%
  select(genus, genus_accuracy) %>%
  distinct() %>%
  summarise(mean(genus_accuracy, na.rm=TRUE))
taxa_guide %>%
  select(family, family_accuracy) %>%
  distinct() %>%
  summarise(mean(family_accuracy, na.rm=TRUE))
taxa_guide %>%
  select(class, class_accuracy) %>%
  distinct() %>%
  summarise(mean(class_accuracy, na.rm=TRUE))
taxa_guide %>%
  select(order, order_accuracy) %>%
  distinct() %>%
  summarise(mean(order_accuracy, na.rm=TRUE))
taxa_guide %>%
  select(phylum, phylum_accuracy) %>%
  distinct() %>%
  summarise(mean(phylum_accuracy, na.rm=TRUE))
```


# now convert into a matrix
```{r}
taxa_knn_mat <- taxa_guide %>% 
  select(species,genus_accuracy, family_accuracy, order_accuracy, class_accuracy, phylum_accuracy) %>%
  column_to_rownames("species") %>%
  as.matrix()

split <- taxa_guide %>%
  filter(species %in% rownames(taxa_knn_mat)) %>%
  arrange(factor(species, levels = rownames(taxa_knn_mat))) %>%
  pull(phylum)
```

## Figure 2D and Supplementary Figure 3A
```{r}
# with na value
cols <- rev(brewer.pal(11, "RdBu"))
col_fun1 <- colorRamp2(seq(0, 1, length.out = 11), cols)
circos.clear()
circos.par(gap.after = c(rep(1,10),20))
circos.heatmap(taxa_knn_mat,
               split = split,
               col=col_fun1,
               track.height = 0.25,
               show.sector.labels = TRUE,
               cluster = FALSE)


lgd = Legend(title = "Accuracy", col_fun = col_fun1)
grid.draw(lgd)

# without NA value
taxa_guide_no_na <- taxa_guide %>% drop_na()
taxa_guide_no_na_mat<- taxa_guide_no_na%>% 
  select(species,genus_accuracy, family_accuracy, order_accuracy, class_accuracy, phylum_accuracy) %>%
  column_to_rownames("species") %>%
  as.matrix()
split <- taxa_guide_no_na %>%
  filter(species %in% rownames(taxa_guide_no_na_mat)) %>%
  arrange(factor(species, levels = rownames(taxa_guide_no_na_mat))) %>%
  pull(phylum)

circos.clear()
circos.heatmap(taxa_guide_no_na_mat,
               split = split,
               col=col_fun1,
               track.height = 0.25,
               show.sector.labels = TRUE,
               cluster = FALSE)


lgd = Legend(title = "Accuracy", col_fun = col_fun1)
grid.draw(lgd)
```




