---
title: "Untitled"
author: "Julia Malnak"
date: "2025-05-28"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# KNN analysis
```{r}
library(tidyverse)
library(RColorBrewer)
library(paletteer)
library(ggrastr)
```


## Supplementary Figure 9: Picking k for KNN
```{r}
# picking for family
molecule_type<- read_tsv("../intermediate_files/family_molecule_type.tsv")  %>% filter(!if_any(everything(), is.na)) %>% distinct()
family_annotations<-read_tsv("../intermediate_files/family_species.tsv") %>% rename(Family=family) %>%
  filter(!is.na(Family)) %>% distinct()
family_shared<- molecule_type %>% inner_join(family_annotations, by="Family")
# phyla annotation
phyla<- read_tsv("../intermediate_files/phyla_species.tsv")
# class annotation
class<- read_tsv("../intermediate_files/class_species.tsv")
# order annotation
order<-read_tsv("../intermediate_files/order_species.tsv")
# genus annotation
genus<-read_tsv("../intermediate_files/genus_species.tsv")
```
### Look at the number k
```{r}
distribution_taxa_plot<- function(df,taxa_level, ylabel)
{
  taxa_sym <- ensym(taxa_level)
  df%>%
    distinct() %>%
    group_by(!!taxa_sym) %>%
    mutate(mems=n()) %>%
    ungroup() %>%
    filter(mems>=2)%>%
    mutate(!!taxa_sym := reorder(!!taxa_sym, mems))%>%
    ggplot(mapping=aes(x=!!taxa_sym))+
    geom_histogram(stat="count", color="white")+
    theme_classic()+
    scale_y_log10()+
    ylab(ylabel)+
    theme(
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      legend.key.height = unit(1.5, "cm"),
       legend.key.width = unit(0.5, "cm"),
      legend.text=element_text(size=7)
    )+
    geom_hline(yintercept = c(3, 5, 11), color="darkblue")

}

```

## Supplementary Figures 9A,D,E,F,G
```{r}
distribution_taxa_plot(family_annotations,Family,"family")
distribution_taxa_plot(phyla, phylum, "Members in Phylum")
distribution_taxa_plot(class,class, "Members in Class")
distribution_taxa_plot(order,order, "Members in Order")
distribution_taxa_plot(genus, genus, "Members in genus")
```

## Supplementary Figure 9B
### Stratify the baltimore classification to see if this is Biased against a single molecular type
```{r}
family_shared %>%
  group_by(Family) %>%
  mutate(mems=n()) %>%
  ungroup() %>%
  filter(mems>=2)%>%
  mutate(Family = reorder(Family, mems))%>%
  ggplot(mapping=aes(x=Family))+
  geom_histogram(stat="count", color="white")+
  facet_wrap(~Molecule_type, scales = "free")+
 # scale_fill_gradientn(
 #   colors=c(rev(brewer.pal(11,"Spectral"))),
 #   limits=c(0,50),
 #   breaks=c(0,3,5,7,11,20,30,40,50, 1000),
 #   name="Number of Members"
 # )+
  theme_classic()+
  scale_y_log10()+
  ylab("Number of Family Members")+
  theme(
    legend.key.height = unit(1.5, "cm"),
     legend.key.width = unit(0.5, "cm"),
    legend.text=element_text(size=7)
  )+
  geom_hline(yintercept = c(3, 5, 11), color="darkblue")

  
```

# look at the median and 60% quantile
```{r}
family_shared %>%
  group_by(Family) %>%
  mutate(mems=n()) %>%
  ungroup() %>%
  filter(mems>=2) %>%
  select(Family,Molecule_type,mems) %>%
  distinct() %>%
  group_by(Molecule_type) %>%
  summarise(med=median(mems))

family_shared %>%
  group_by(Family) %>%
  mutate(mems=n()) %>%
  ungroup() %>%
  filter(mems>=2)%>%
  select(Family,Molecule_type,mems) %>%
  distinct() %>%
  group_by(Molecule_type) %>%
  summarise(med=quantile(mems, 0.75))

  
```

## Main figure: accuracy of KNN performance within the family
```{r}
# this is for the symmetric jaccard
knn_3<- read_csv("../intermediate_files/results_knn/knn_result_k_3_structure.csv")
knn_5<-read_csv("../intermediate_files/results_knn/knn_result_k_5_structure.csv")
knn_11<-read_csv("../intermediate_files/results_knn/knn_result_k_11_structure.csv")

```

```{r}
accuracy<-function(knn_df)
{
  knn_acc<- knn_df %>% group_by(known_family) %>%
  summarise(correct=sum(known_family==predicted_family),
         overall=n()) %>%
  mutate(accuracy=correct/overall) %>%
  mutate(known_family = reorder(known_family, -accuracy))
  
  return(knn_acc)
}
```

```{r}
knn_acc_3<- accuracy(knn_3)
knn_acc_5<- accuracy(knn_5)
knn_acc_11<- accuracy(knn_11)
```

```{r}
plot_accuracy<-function(knn)
{
  knn %>%
    ggplot(mapping=aes(x=known_family, y=accuracy))+
    rasterise(
    geom_bar(stat = "identity", color = "white", linewidth = 0.25),
    dpi = 300
  )+
    theme_classic()+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size=10),
         axis.text.y = element_text(angle = 90, hjust = 0.5))+
    ylab("Accuracy")+
    xlab("Family")
}
```
## Figure 2A, Suuplementary Figure 2A,B
```{r}
plot_accuracy(knn_acc_3)
plot_accuracy(knn_acc_5)
plot_accuracy(knn_acc_11)

```


## Supplementary Figure 9C: Performance across different values of K
```{r}
knn_acc_3<-knn_acc_3 %>% mutate(k=factor(3))
knn_acc_5<- knn_acc_5 %>% mutate(k=factor(5))
knn_acc_11 <- knn_acc_11 %>% mutate(k=factor(11))
```

```{r}
lg_knn<- knn_acc_3 %>% bind_rows(knn_acc_5) %>% bind_rows(knn_acc_11) %>% rename(family=known_family)
lg_knn %>%
  ggplot(mapping = aes(x=k, y=accuracy, group=k, color=k))+
  geom_violin(fill=NA, linewidth=1, outliers = FALSE)+
  geom_jitter(alpha=0.6, size=1.75, stroke=0.7)+
  theme_classic()+
  xlab("Hyperparameter K")+
  ylab("Accuracy")+
  scale_color_paletteer_d("PrettyCols::Bright")+
  theme(legend.position="none",
        axis.text.x = element_text( size=12))
```

### Significance testing for differences of hyperparameter
```{r}
library(lme4)
# building a linear mixed effects model
# accuracy ~ k + (1|family), family is the random effect because not independent.
k_mod<-lmer(accuracy~k + (1|family), data=lg_knn)
drop1(k_mod,test="Chisq")
```


## Supplementary: Performance across average family size

```{r}
family_size<-read_csv("../intermediate_files/nomburg_files/ncbi_viruses_genome_lengths.csv") %>%
  select(Family,Molecule_type,Species, Length ) %>%
  filter(!if_any(everything(), is.na)) %>%
  group_by(Family) %>%
  mutate(avg_length=mean(Length, na.rm=TRUE)) %>%
  select(Family, Molecule_type,avg_length) %>%
  rename(family=Family)  %>%
  distinct() %>%
  left_join(lg_knn, by="family") %>%
  filter(!if_any(everything(), is.na)) %>%
  mutate(avg_length=avg_length/1000)
```
### Supplementary Figure 3a, Size vs Sensitivity
```{r}
family_size %>%
  filter(k==11) %>%
  ggplot(mapping=aes(x=avg_length, y=accuracy))+
  geom_point(size=2)+
  scale_x_log10()+
  scale_color_paletteer_d("PrettyCols::Bright")+
  theme_classic()+
  xlab("Length in kb")+
  ylab("Family Accuracy")#+
  #facet_wrap(~Molecule_type, scales="free")
```

```{r}
## spearman correlation calculation
# correlation
cor(family_size$avg_length, family_size$accuracy, method="spearman")
# p value
cor.test(family_size$avg_length, family_size$accuracy, method="spearman")
```

### Number of features vs accuracy
```{r}
count_df<- read_csv("../intermediate_files/structure_df.csv") 
```

```{r}
count_df<-count_df %>% pivot_longer(cols=-cluster_ID,names_to="species", values_to = "Present" ) 
count_df<- count_df %>% inner_join(family_annotations, by="species") %>% 
  select(-species) %>%
  group_by(Family, cluster_ID) %>%
  mutate(presence=ifelse(sum(Present)>0, 1, 0)) %>%
  ungroup() %>%
  select(Family, cluster_ID, presence)
# now count the number of features
count_df<- count_df %>%
  distinct() %>% 
  select(-cluster_ID) %>%
  group_by(Family) %>%
  summarise(features=sum(presence)) %>%
  left_join(knn_acc_11, by=c("Family"="known_family")) %>% 
  drop_na()
  
```
## Supplementary Figure 3B
```{r}
options(scipen=999)
# plotting features vs accuracy
count_df %>%
  #mutate(log10_feat=log10(features)) %>%
  mutate(bin=cut(features, seq(min(features),max(features+10),10), right = FALSE)) %>%
  ggplot(mapping=aes(x=bin, y=accuracy))+
  stat_summary(fun=mean, geom="bar", fill="skyblue")+
  geom_jitter(width=0.2, size=2, alpha=0.5, shape=16)+
  theme_classic()+
  geom_hline(aes(yintercept=0.5), color="red")+
  xlab("Binned Number of Features")+
  ylab("Accuracy")+
  theme(axis.text.x=element_text(angle = 90, vjust=0.5, hjust=1))
```

```{r}
## spearman correlation calculation
# correlation
cor(count_df$features, count_df$accuracy, method="spearman")
# p value
cor.test(count_df$features, count_df$accuracy, method="spearman")
```


## Add in additional analysis of the knn results from the subsampled structures
### Supplementary Figure 2C
```{r}
baseline<-read_csv("../intermediate_files/results_knn/knn_result_k_11_structure.csv") %>%
  rename("known_taxa"="known_family") %>%
  rename("predicted_taxa"="predicted_family")
plddt_50<-read_csv("../intermediate_files/high_quality_structure_results/knn_plddt_50.csv")
plddt_70<-read_csv("../intermediate_files/high_quality_structure_results/knn_plddt_70.csv")
```

# different accuracy function
```{r}
mod_accuracy<-function(knn_df)
{
  knn_acc<- knn_df %>% group_by(known_taxa) %>%
  mutate(correct=sum(known_taxa==predicted_taxa),
         overall=n()) %>%
  mutate(accuracy=correct/overall) %>%
  mutate(known_taxa = reorder(known_taxa, -accuracy)) %>%
    ungroup() %>%
    select(species,accuracy)
  
  return(knn_acc)
}
```

# get the accuracy for each and then combine into the large plot
```{r}
# filtering out 
baseline<-mod_accuracy(baseline) %>% mutate(plddt="baseline")
plddt_50<-mod_accuracy(plddt_50) %>% mutate(plddt="50") 
plddt_70<-mod_accuracy(plddt_70) %>% mutate(plddt="70") 

# put all together
knn_all_plddt<- baseline %>% 
  rbind(plddt_50) %>%
  rbind(plddt_70) 
```

```{r}
# join with family and look at family level accuracy
plddt_accuracy<-knn_all_plddt %>%
  left_join(family_annotations, by="species") %>%
  select(-species) %>%
  distinct()
# filter for families only in baseline
baseline_fam <- plddt_accuracy %>% filter(plddt=="baseline") %>% pull(Family)
plddt_accuracy <- plddt_accuracy %>% filter(Family %in% baseline_fam)
```

# plot the plddt accuracy 
```{r}
pal <- paletteer_d("khroma::BuRd")
continuous_pal <- colorRampPalette(pal)(100)
# plotting results
plddt_accuracy %>%
  mutate(plddt=factor(plddt, levels=c("baseline", "50", "70"))) %>%
  ggplot(mapping=aes(x=plddt, y=Family, fill=accuracy))+
  geom_tile(color="white", linewidth=0.1)+
  scale_fill_gradientn(colors=continuous_pal)+ # changing the fill palette
  theme_classic()+
  xlab("pLDDT of structures")+
  theme(axis.title.y = element_blank(), 
        plot.title = element_text(color = "black"),
        axis.title.x = element_text(color = "black"),
        axis.text.x = element_text(color = "black", angle=90),
        axis.text.y = element_text(color = "black"),
        legend.text = element_text(color = "black"),
        legend.title = element_text(color = "black"),
        legend.key = element_rect(colour = NA))+
  coord_fixed(ratio=0.5) # making the plot smaller
```

Calculate the average for each strata
```{r}
plddt_accuracy %>% filter(plddt=="baseline") %>% pull(accuracy) %>% mean()
plddt_accuracy %>% filter(plddt=="baseline") %>% pull(accuracy) %>% var()
plddt_accuracy %>% filter(plddt=="50") %>% pull(accuracy) %>% mean()
plddt_accuracy %>% filter(plddt=="50") %>% pull(accuracy) %>% var()
plddt_accuracy %>% filter(plddt=="70") %>% pull(accuracy) %>% mean()
plddt_accuracy %>% filter(plddt=="70") %>% pull(accuracy) %>% var()
```

```{r}
virus_b<- plddt_accuracy %>% filter(plddt=="baseline") %>% pull(Family)
virus_50<- plddt_accuracy %>% filter(plddt=="50") %>% pull(Family)
virus_70<-plddt_accuracy %>% filter(plddt=="70") %>% pull(Family)
virus_overlap <- virus_b %>% intersect(virus_50) %>% intersect(virus_70) 
plddt_accuracy<- plddt_accuracy %>% filter(Family %in% virus_overlap)
```

```{r}
plddt_accuracy %>% filter(plddt=="baseline") %>% pull(accuracy) %>% mean()
plddt_accuracy %>% filter(plddt=="baseline") %>% pull(accuracy) %>% var()
plddt_accuracy %>% filter(plddt=="50") %>% pull(accuracy) %>% mean()
plddt_accuracy %>% filter(plddt=="50") %>% pull(accuracy) %>% var()
plddt_accuracy %>% filter(plddt=="70") %>% pull(accuracy) %>% mean()
plddt_accuracy %>% filter(plddt=="70") %>% pull(accuracy) %>% var()

```

