---
title: "robustness_analysis"
author: "Julia Malnak"
date: "2025-07-21"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Robustness analysis

## Loading in the libraries and data
```{r}
library(tidyverse)
library(paletteer)
library(UpSetR)
```

```{r}
# loading in the data
# this is the baseline performance
feat_100<-read_csv("../intermediate_files/results_knn/knn_result_k_11_structure.csv") %>%
  mutate(iteration=0) # to make compatible for acccuracy function
# this is the performance of the bootstrap
# feat_x where x is the percentage of the features removed
feat_5<-read_csv("../intermediate_files/robustness_knn/5_percent/all_iterations_results.csv", col_names = c("species", "known_family", "predicted_family", "score", "iteration"))
feat_10<- read_csv("../intermediate_files/robustness_knn/10_percent/all_iterations_results.csv", col_names = c("species", "known_family", "predicted_family", "score", "iteration"))
feat_20<-read_csv("../intermediate_files/robustness_knn/20_percent/all_iterations_results.csv", col_names = c("species", "known_family", "predicted_family", "score", "iteration"))
feat_50<-read_csv("../intermediate_files/robustness_knn/50_percent/all_iterations_results.csv", col_names = c("species", "known_family", "predicted_family", "score", "iteration"))
feat_75<-read_csv("../intermediate_files/robustness_knn/75_percent/all_iterations_results.csv", col_names = c("species", "known_family", "predicted_family", "score", "iteration"))
feat_90<-read_csv("../intermediate_files/robustness_knn/90_percent/all_iterations_results.csv", col_names = c("species", "known_family", "predicted_family", "score", "iteration"))
family_annotations<-read_tsv("../intermediate_files/family_species.tsv") %>% drop_na()

```
Now to analyze for sensitivity
```{r}
# this is actually the sensitivity function but named improperly
accuracy<-function(knn_df)
{
  
  knn_acc<- knn_df %>% group_by(iteration,known_family) %>% # group family for each iteration
  summarise(correct=sum(known_family==predicted_family), # get the number correct
         overall=n()) %>% # number of members in the family
  mutate(accuracy=correct/overall) %>% # sensitivity is the amount correct
  mutate(known_family = reorder(known_family, -accuracy)) %>% 
  ungroup()  %>%  # ungrouping from the iterations
  group_by(known_family) %>% # group by family
  mutate(mean_accuracy=mean(accuracy)) %>% # get average performance across iterations
  ungroup()
  
  return(knn_acc)
}
```

```{r}
# calculate the sensitivity and add the robustness as a feature to the df
feat_100_results<- accuracy(feat_100) %>% mutate(removed=0)
feat_5_results<-accuracy(feat_5) %>% mutate(removed=5)
feat_10_results<- accuracy(feat_10) %>% mutate(removed=10)
feat_20_results<- accuracy(feat_20) %>% mutate(removed=20)
feat_50_results<- accuracy(feat_50) %>% mutate(removed=50)
feat_75_results<- accuracy(feat_75) %>% mutate(removed=75)
feat_90_results<- accuracy(feat_90) %>% mutate(removed=90)
```
Plot the mean accuracy and the number of features removed
```{r}
# put together all the results into one df
all_results<- feat_100_results %>% 
  rbind(feat_5_results) %>%
  rbind(feat_10_results) %>%
  rbind(feat_20_results) %>%
  rbind(feat_50_results) %>%
  rbind(feat_75_results) %>%
  rbind(feat_90_results)
 
```
## Figure 2B
```{r}
# add the diverging palette from paletteer package this is a blue-red palette
pal <- paletteer_d("khroma::BuRd")
continuous_pal <- colorRampPalette(pal)(100)

# plot the average sensitivity at each stage
all_results %>%
  select(known_family, mean_accuracy, removed) %>%
  distinct() %>%
  mutate(removed=factor(removed)) %>%
  ggplot(mapping=aes(x=removed, y=known_family, fill=mean_accuracy))+
  geom_tile(color="white", linewidth=0.1)+
  scale_fill_gradientn(colors=continuous_pal)+ # changing the fill palette
  theme_classic()+
  xlab("% Removed")+
  theme(axis.title.y = element_blank(), 
        plot.title = element_text(color = "black"),
        axis.title.x = element_text(color = "black"),
        axis.text.x = element_text(color = "black"),
        axis.text.y = element_text(color = "black"),
        legend.text = element_text(color = "black"),
        legend.title = element_text(color = "black"),
        legend.key = element_rect(colour = NA))+
  coord_fixed(ratio=0.5) # making the plot smaller
```

##Calculate statistics on the drop of sensitivity between points
```{r}
all_results %>% 
  select(known_family, mean_accuracy, removed) %>%
  distinct() %>%
  pivot_wider(values_from = mean_accuracy, names_from = removed) %>%
  mutate(
    drop_5=`0`-`5`,
    drop_10=`0`-`10`,
    drop_20=`0`-`20`,
    drop_50=`0`-`50`,
    drop_75=`0`-`50`,
    drop_90=`0`-`90`) %>%
  select(-known_family) %>% 
  summarise(across(everything(), ~ mean(.x, na.rm = TRUE)))
```


# Look at the results across different removals
Filter for viral families that did see a drop in performance!
Look at each time there is a drop count the protein that was involved!

```{r}
# add in the feature list
features_5<-read_csv("../intermediate_files/robustness_knn/5_percent/feat_removal.csv", col_names = c("features", "iteration"))
features_10<-read_csv("../intermediate_files/robustness_knn/10_percent/feat_removal.csv", col_names = c("features", "iteration"))
features_20<- read_csv("../intermediate_files/robustness_knn/20_percent/feat_removal.csv", col_names = c("features", "iteration"))
features_50<-read_csv("../intermediate_files/robustness_knn/50_percent/feat_removal.csv", col_names = c("features", "iteration"))
features_75<-read_csv("../intermediate_files/robustness_knn/75_percent/feat_removal.csv", col_names = c("features", "iteration"))
features_90<-read_csv("../intermediate_files/robustness_knn/90_percent/feat_removal.csv", col_names = c("features", "iteration"))
all_features<-read_csv("../intermediate_files/cluster_mems.csv")

```

Make a function to look for boost in performance and count the cluster which was removed.
```{r}
# merge the default performance
better_performance<- function(df, benchmark_performance, features, filter)
{
  # normalize by the number by which a feature is selected
  features<- features %>% 
    group_by(features) %>%
    mutate(norm=n()) %>%
    ungroup()
  
  subset<-df %>% 
    select(iteration, known_family, accuracy) %>% # iteration of result, family, accuracy
    left_join(benchmark_performance, by="known_family") %>% # add the benchmark performance
    filter(accuracy > benchmark_accuracy) %>%  # only keep rows where better
    select(-accuracy, -benchmark_accuracy) %>% # remove the accuracy
    left_join(features, by="iteration") %>% # add in the features
    select(iteration,features, known_family, norm) %>% # select the norm, features, and iteration
    distinct() %>%
    select(-iteration) %>% # remove iteration not needed
    group_by(known_family,features) %>% # count the number of times that each feature is important
    mutate(count=n(),
           norm_count=count/norm) %>%
    ungroup() %>%
    mutate(z_score_norm_count = (norm_count - mean(norm_count, na.rm = TRUE)) /  # z score across all features
                                sd(norm_count, na.rm = TRUE)) %>% 
    filter(count>filter) %>% 
    filter(z_score_norm_count > 1.645 | z_score_norm_count < -1.645) %>% # filter at the top 5 percentiles
    arrange(desc(norm_count)) 
    
  return(subset)
}

```


```{r}
# preprocess perfect performance
benchmark_performance<- feat_100_results %>% select(known_family,accuracy) %>%
  rename("benchmark_accuracy"=accuracy) %>% distinct()
feat_5_clusts<-better_performance(feat_5_results, benchmark_performance, features_5,3) 
feat_10_clusts<-better_performance(feat_10_results, benchmark_performance, features_10, 5) 
feat_20_clusts<-better_performance(feat_20_results, benchmark_performance, features_20, 10) 
feat_50_clusts<-better_performance(feat_50_results, benchmark_performance, features_50, 25) 
feat_75_clusts<-better_performance(feat_75_results, benchmark_performance, features_75, 38) 
feat_90_clusts<-better_performance(feat_90_results, benchmark_performance, features_90, 45) 


# look to see if there is any overlap in the features (upset plot here)
# make into named list of features
clusts<-list(removed_5=feat_5_clusts$features, removed_10=feat_10_clusts$features,
             removed_20=feat_20_clusts$features, 
             removed_50=feat_50_clusts$features,
             removed_75=feat_75_clusts$features,
             removed_90=feat_90_clusts$features)
# plot the overlaps
upset(fromList(clusts), order.by="freq", point.size = 2.5, line.size = 1.5)


```

Now going to look at what the clusters are that improve performance
```{r}
improve_features <- intersect(feat_5_clusts$features, feat_10_clusts$features) %>%
  intersect(feat_20_clusts$features) 

# these features must be contained by the virus that the family is misclassified for at baseline
baseline<-read_csv("../intermediate_files/results_knn/knn_result_k_11_structure.csv") %>% filter(known_family!=predicted_family)
count_df<- read_csv("../intermediate_files/structure_df.csv") 
```

```{r}
# create df of features available within the family
merged_count_df<-count_df %>% pivot_longer(cols=-cluster_ID,names_to="species", values_to = "Present" ) %>%  filter(cluster_ID %in% improve_features)

merged_count_df<- merged_count_df %>% inner_join(family_annotations, by="species") %>% 
  select(-species) %>%
  group_by(family, cluster_ID) %>%
  mutate(presence=ifelse(sum(Present)>0, 1, 0)) %>%
  ungroup() %>%
  select(family, cluster_ID, presence) 
```

```{r}
baseline <- baseline %>% select(-species) %>% distinct() %>% left_join(merged_count_df, by=c("known_family"="family")) %>% 
  filter(presence==1)
# rename so they don't get .x and .y after
merged_count_df <- merged_count_df %>% rename(class_presence=presence)
baseline <- baseline %>% left_join(merged_count_df, by=c("predicted_family"="family", "cluster_ID")) %>% distinct()
# only filter where shared
baseline<- baseline %>% filter(presence==class_presence)
# this df includes clusters found in viruses that improve performance when removed.

```

# Now visualizing these clusters of interest
```{r}
# merge the default performance
better_performance_features<- function(df, benchmark_performance, features, filter)
{
  # normalize by the number by which a feature is selected
  features<- features %>% 
    group_by(features) %>%
    mutate(norm=n()) %>%
    ungroup()
  
  subset<-df %>% 
    select(iteration, known_family, accuracy) %>% # iteration of result, family, accuracy
    left_join(benchmark_performance, by="known_family") %>% # add the benchmark performance
    filter(accuracy > benchmark_accuracy) %>%  # only keep rows where better
    mutate(change_performance=accuracy-benchmark_accuracy) %>%
    select(-accuracy, -benchmark_accuracy) %>% # remove the accuracy
    left_join(features, by="iteration") %>% # add in the features
    select(iteration,features, known_family, norm, change_performance) %>% # select the norm, features, and iteration
    distinct() %>%
    select(-iteration) %>% # remove iteration not needed
    group_by(known_family,features) %>% # count the number of times that each feature is important
    mutate(count=n(),
           norm_count=count/norm) %>%
    ungroup() %>%
    mutate(z_score_norm_count = (norm_count - mean(norm_count, na.rm = TRUE)) /  # z score across all features
                                sd(norm_count, na.rm = TRUE)) %>% 
    filter(count>filter) %>%  # this will get rid of noise when removed once
    distinct() %>%
    arrange(desc(norm_count)) 
    
  return(subset)
}
```

```{r}
feat_5_clusts_feat<-better_performance_features(feat_5_results, benchmark_performance, features_5,3)
feat_10_clusts_feat<-better_performance_features(feat_10_results, benchmark_performance, features_10, 5) 
feat_20_clusts_feat<-better_performance_features(feat_20_results, benchmark_performance, features_20, 10) 
feat_50_clusts_feat<-better_performance_features(feat_50_results, benchmark_performance, features_50, 25) 
feat_75_clusts_feat<-better_performance_features(feat_75_results, benchmark_performance, features_75, 38) 
feat_90_clusts_feat<-better_performance_features(feat_90_results, benchmark_performance, features_90, 45) 
```


Now looking for clusters in which removing results a decrease in performance
```{r}
# merge the default performance
worse_performance<- function(df, benchmark_performance, features, filter)
{
  # normalize by the number by which a feature is selected
  features<- features %>% 
    group_by(features) %>%
    mutate(norm=n()) %>%
    ungroup()
  
  subset<-df %>% 
    select(iteration, known_family, accuracy) %>% # iteration of result, family, accuracy
    left_join(benchmark_performance, by="known_family") %>% # add the benchmark performance
    filter(accuracy < benchmark_accuracy) %>%  # only keep rows where worse
    select(-accuracy, -benchmark_accuracy) %>% # remove the accuracy
    left_join(features, by="iteration") %>% # add in the features
    select(iteration,features, known_family, norm) %>% # select the norm, features, and iteration
    distinct() %>%
    select(-iteration) %>% # remove iteration not needed
    group_by(known_family,features) %>% # count the number of times that each feature is important
    mutate(count=n(),
           norm_count=count/norm) %>%
    ungroup() %>%
    filter(count>filter) %>%  # this will get rid of noise when removed once
    distinct() %>%
    arrange(desc(norm_count)) %>%
    filter(norm_count > 0.5) # must at least hurt performance half the time it gets removed
    
  return(subset)
}

```
### Supplementary Figure 2E
```{r}
feat_5_clusts_worse<-worse_performance(feat_5_results, benchmark_performance, features_5,2)
feat_10_clusts_worse<-worse_performance(feat_10_results, benchmark_performance, features_10,5)
feat_20_clusts_worse<-worse_performance(feat_20_results, benchmark_performance, features_20, 10)
feat_50_clusts_worse<-worse_performance(feat_50_results, benchmark_performance, features_50, 25)
feat_75_clusts_worse<-worse_performance(feat_75_results, benchmark_performance, features_75, 37)
feat_90_clusts_worse<-worse_performance(feat_90_results, benchmark_performance, features_90, 45)
# show how much overlap there is
clusts<-list(removed_5=feat_5_clusts_worse$features, 
             removed_10=feat_10_clusts_worse$features,
             removed_20=feat_20_clusts_worse$features, 
             removed_50=feat_50_clusts_worse$features,
             removed_75=feat_75_clusts_worse$features,
             removed_90=feat_90_clusts_worse$features)
# plot the overlaps
common_features <- Reduce(intersect, clusts)
upset(fromList(clusts), order.by="freq", point.size = 2.5, line.size = 1.5)
```

# There is a giant overlap
So I will merge the features by the family and the features
```{r}
# filter for relevant columns
worse_5<- feat_5_clusts_worse %>% select(known_family,features, norm_count) %>% rename(count_5=norm_count)
worse_10<- feat_10_clusts_worse %>% select(known_family,features, norm_count)%>% rename(count_10=norm_count)
worse_20<- feat_20_clusts_worse %>% select(known_family,features, norm_count)%>% rename(count_20=norm_count)
worse_50<- feat_50_clusts_worse %>% select(known_family,features, norm_count)%>% rename(count_50=norm_count)
worse_75<- feat_75_clusts_worse %>% select(known_family,features, norm_count)%>% rename(count_75=norm_count)
worse_90<- feat_90_clusts_worse %>% select(known_family,features, norm_count)%>% rename(count_90=norm_count)
# merge all features of the same family
worse_performance<- worse_5 %>% inner_join(worse_10, by=c("known_family", "features")) %>%
  inner_join(worse_20, by=c("known_family", "features")) %>% 
  inner_join(worse_50, by=c("known_family", "features")) %>%
  inner_join(worse_75, by=c("known_family", "features")) %>%
  inner_join(worse_90, by=c("known_family", "features"))
# I'm also going to filter for families and features where the feature is represented because we are getting many unexplainable features
merged_count_df<-count_df %>% 
  pivot_longer(cols=-cluster_ID,names_to="species", values_to = "Present") %>%
  filter(Present==1) %>%
  inner_join(family_annotations, by="species") %>% 
  select(-species) %>%
  group_by(family, cluster_ID) %>%
  mutate(presence=ifelse(sum(Present)>0, 1, 0)) %>%
  ungroup() %>%
  select(family, cluster_ID, presence) %>%
  drop_na() %>% 
  distinct() %>% 
  filter(presence==1) 
```

```{r}
# now I'm going to right merge the df to only look at features in that family
worse_performance<- merged_count_df %>% select(-presence) %>% inner_join(worse_performance, by=c("family"="known_family", "cluster_ID"="features"))
```

## How many clusters are worse for performance and how many of them are family-specific?
```{r}
unique_features<- merged_count_df %>%
  group_by(cluster_ID) %>%
  mutate(unique=n()) %>%
  ungroup() %>%
  filter(unique==1) %>% 
  pull(cluster_ID)
worse_performance <- worse_performance %>%
  mutate(unique=ifelse(cluster_ID %in% unique_features, "Specific", "Non-specific"))
```

# Main Figure 2C
```{r}
# plot the number of features that hurt performance and how many are family specific
worse_performance %>%
  group_by(family) %>%
  mutate(counting=n()) %>%
  ungroup() %>%
  mutate(family=reorder(family,-counting)) %>%
  ggplot(mapping=aes(fill=unique, x=family))+
  geom_bar(stat="count")+
  scale_fill_paletteer_d("nationalparkcolors::CraterLake")+
  theme_classic()+
  theme(axis.text.x=element_text(angle = 90, vjust=0.5, hjust=1))+
  ylab("Features")+
  xlab("Family")+
  labs("Family specificity")
```

## Testing the significance of the appearance of more family specific clusters
```{r}
# 17,447 total
# 14,261 are unique
# 140 are sampled after filtering
# 110 are family specific
phyper(110,14698,17447-14698, 140, lower.tail = FALSE)
```

## Supplementary Figure 2D: Is robustness correlated with the larger vectors?
```{r}
options(scipen=9999)
# count the number of features per family
family_count<- merged_count_df %>%
  group_by(family) %>%
  summarise(features=n())
# look at the robustness by the accuracy
all_results %>%
  select(known_family, mean_accuracy, removed) %>%
  left_join(family_count, by=c("known_family"="family")) %>%
  mutate(removed=as.factor(removed)) %>%
  ggplot(mapping=aes(x=features, y=mean_accuracy, color=removed))+
  rasterise(geom_point(), dpi=300)+
  rasterise(geom_line(), dpi=300)+
  scale_x_log10()+
  theme_classic()+
  facet_wrap(~removed)+
   scale_color_paletteer_d("ggthemes::Jewel_Bright")

```

## spearman correlation between features and accuracy
```{r}
spear_df<- all_results %>%
  select(known_family, mean_accuracy, removed) %>%
  left_join(family_count, by=c("known_family"="family"))
# 5 % removed
correlation_calc<- function(removal)
{
print(paste("results for", removal, "."))
spear<-spear_df %>% filter(removed==removal) 
cor.test(spear$mean_accuracy,spear$features,method="spearman")
}
correlation_calc(5)
correlation_calc(10)
correlation_calc(20)
correlation_calc(50)
correlation_calc(75)
correlation_calc(90)
```

