---
title: "conserved_gene_analysis"
author: "Julia Malnak"
date: "2025-07-03"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(paletteer)
library(pheatmap)
library(reshape2)
```

# Shared similarity analysis
Loading and preprocessing the data
```{r}
str_sim<- read_csv("../intermediate_files/species_jaccard.csv") %>%
  select(-union, -intersection) %>%
  rename(jaccard_similarity=`jaccard index`) %>%
  filter(taxon1!="Staphylococcus_aureus") %>%
  filter(taxon2!="Staphylococcus_aureus")
family_anno<-read_tsv("../intermediate_files/family_species.tsv") %>% drop_na() %>% distinct() %>%
  group_by(family) %>%
  mutate(mems=n()) %>%
  ungroup
```

Join the family with the df.
```{r}
str_sim<- str_sim %>% inner_join(family_anno, by=c("taxon1"="species"))%>% 
  rename(family1=family) %>%
  inner_join(family_anno, by=c("taxon2"="species")) %>% 
  rename(family2=family) 
```



## Looking for which nearest neighbors are not always from the within family. 
This is essentially a heatmap of counts of who the nearest neighbor is (top 1)
```{r}
str_sim_grouped<- str_sim %>%
  group_by(taxon1) %>%
  filter(min_rank(desc(jaccard_similarity)) <= 1) %>%
  ungroup()
str_sim_grouped_no_match<- str_sim_grouped %>%
  filter(jaccard_similarity!=0)
# now let's only include things that has a nearest neighbor that's not in its family
viruses_of_interest<- str_sim_grouped_no_match %>%
  filter(family1!=family2) %>%
  pull(family1)
virus_of_interest<- str_sim_grouped_no_match %>%
  filter(family1!=family2) %>%
  select(taxon1, taxon2) %>%
  as.list() %>%
  unlist() %>%
  unique()
# sub df for figure
counts_cross_match <- str_sim_grouped_no_match %>%
  filter(family1 %in% viruses_of_interest) %>%
  select(family1,family2) %>%
  group_by(family1,family2) %>%
  summarise(counts=n()) %>%
  ungroup()
```
## Figure 3A
```{r}
family_size<- family_anno %>% select(family, mems) %>% distinct()
### normalization of the dataframe
prep_mat<- counts_cross_match %>%
  mutate(counts=log10(counts+1)) %>%
  pivot_wider(names_from = family2, values_from = counts, values_fill = 0) %>%
  column_to_rownames("family1") %>%
  as.matrix()

colors_continuous <- colorRampPalette(paletteer_d("MoMAColors::Althoff"))(100)

# Make sure 0 maps to white and rest uses the gradient
color_vector <- c("white", colors_continuous)


# heatmap function
pheatmap(prep_mat,
         color=color_vector, 
         legend = TRUE)
```


Now look at the 11 most close neighbors!!
```{r}
# eliminate families without 11 neighbors

str_sim_grouped_11<- str_sim %>%
  filter(mems.x >11) %>%
  group_by(taxon1) %>%
  filter(min_rank(desc(jaccard_similarity)) <= 11) %>%
  ungroup()
# now going to calculate the cases in which a misclassification would be mades
str_sim_grouped_11 <- str_sim_grouped_11 %>%
  filter(jaccard_similarity!=0) %>%
  group_by(taxon1) %>%
  mutate(total=n()) %>%
  ungroup() %>%
  group_by(taxon1, family2) %>%
  mutate(class=n()) %>%
  mutate(majority_rules=class/total) %>%
  ungroup() %>%
  group_by(taxon1) %>%
  filter(min_rank(desc(majority_rules)) <= 1) %>%
  ungroup() %>%
  select(taxon1,family1, family2) %>%
  distinct() %>%
  select(family1,family2) %>%
  group_by(family1,family2) %>%
  mutate(counts=n()) %>%
  ungroup() %>%
  distinct()
# now only look at families which are misclassified
families_interest<- str_sim_grouped_11 %>%
  filter(family1!=family2) %>%
  pull(family1)
#only include viruses that are misclassified at least once
str_sim_grouped_11 <- str_sim_grouped_11 %>%
  filter(family1 %in% families_interest)
```


## Supplementary Figure 4A
```{r}
prep_mat<- str_sim_grouped_11 %>%
  mutate(counts = log10(counts + 1)) %>%
  pivot_wider(names_from = family2, values_from = counts, values_fill = list(counts = 0)) %>%
  column_to_rownames("family1") %>%
  as.matrix()
# heatmap function

# making palette
my_palette <- paletteer_d("MoMAColors::Ernst") %>% unname()
color_gradient <- colorRampPalette(c("white", my_palette))(100)
min_val <- min(prep_mat, na.rm = TRUE)
max_val <- max(prep_mat, na.rm = TRUE)
breaks <- seq(0, max_val, length.out = 101)
#plot
pheatmap(prep_mat,
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         color = color_gradient,
         breaks = breaks, cellwidth = 8,
         cellheight = 9)

```


What are these shared genes between families?
I'm going to import some count results because the matrix is too large to handle in R.

I'm going to subset by genes that are interesting in the other figure!
```{r}
species_count_df<- read_csv("../intermediate_files/structure_df.csv")
```

## Looking at the genes shared between a subset of viruses in the previous heatmap
```{r}
shared_subset<- species_count_df %>%
  select(all_of(c("cluster_ID",virus_of_interest))) %>%
  rowwise() %>%
  mutate(total_count=sum(c_across(-cluster_ID))) %>%
  filter(total_count > 1) 
```

## Figure 3B
```{r}
# plotting results
# prepping the df
shared_subset_mat <- shared_subset %>%
  select(-total_count) %>%
   column_to_rownames("cluster_ID") %>%
  as.matrix() %>%
  t()
order_viruses<- rownames(shared_subset_mat)
shared_subset_viruses <- tibble(species=rownames(shared_subset_mat))
# adding the annotations
family_annotation<- family_anno %>% 
  right_join(shared_subset_viruses, by="species")%>%
  distinct() %>%
  arrange(factor(species, levels = order_viruses)) %>%
  arrange(family)
# order the heatmap in the same way?
shared_subset_mat <- shared_subset_mat[family_annotation$species, ]
## adding the 
family_annotation<- family_annotation %>%
  select(family) %>%
  as.data.frame() %>%
  mutate(family=factor(family))
rownames(family_annotation)<- rownames(shared_subset_mat)
family_levels <- levels(family_annotation$family)
# adding the custom colors
family_col<- c(paletteer_d("palettetown::bulbasaur"),paletteer_d("palettetown::gloom"))
ann_cols<- list(family = setNames(family_col[1:length(family_levels)], family_levels))
# plotting the results
pheatmap(shared_subset_mat,
         col=paletteer_d("soilpalettes::crait"),
         show_rownames = F,
         annotation_row = family_annotation,
         cluster_rows = F,
         cellwidth=8,
          cellheight=0.7, 
         annotation_colors = ann_cols)
```

# Figure 3C
## Add a Corrplot to see if the there are variables that are correlated in the heatmap?
```{r}

library(corrplot)
# correlation
correlation<-cor(shared_subset_mat, use="pair")
# pvalue mat using fisher's exact
c<-rownames(correlation)
p_mat<-matrix(NA, nrow=length(c), ncol=length(c), dimnames=list(c,c))

p_val_list<-list()
index_list<-list()
count<-1
for (i in seq_along(c)) # loop through and calculate
{
  for (j in seq_along(c))
  {
    tab<-table(shared_subset_mat[,i], shared_subset_mat[,j])
    p<- fisher.test(tab)$p.value
    if (i!=j) # don't include self comparisions
    {
      p_val_list[[count]]=p
      index_list[[count]]=c(i,j)
      count<- count+1
    }
  }
}

# calculate the adjusted p-values
p_vals<-unlist(p_val_list)
adj_p<-p.adjust(p_vals, method="fdr")

# fill in the matrix with the adjusted p-values
for (idx in seq_along(index_list))
{
  i<-index_list[[idx]][1]
  j<-index_list[[idx]][2]
  p_mat[i,j]<-adj_p[idx]
  p_mat[j,i]<-adj_p[idx]
}
# fill in NA values for self comparision
diag(p_mat)<-1
# transform for sizing
p_mat<--log10(p_mat)

## visualization of correlation
p<-melt(p_mat) %>% rename("logq"="value")
c<-melt(correlation) %>% rename("coeff"="value")

## make the color palette
pal<-colorRampPalette(paletteer_d("colorBlindness::Blue2Orange10Steps"))(100)

df <- p %>% inner_join(c, by=c("Var1","Var2")) %>%
  filter(Var1 < Var2) %>%
  mutate(Var1=as.factor(Var1)) %>%
  mutate(Var2=as.factor(Var2))
ggplot(df, aes(x = Var1, y = Var2)) +
  geom_tile(fill="white", color = "black") +  # boxes
  geom_point(aes(fill=coeff, size = logq), shape = 21) + # points sized by significance
  scale_fill_gradientn(colors=pal) +
  scale_size(range = c(0.5, 6)) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    axis.title = element_blank()
  ) +
  labs(fill = "Phi-Correlation", size = "-log(q-value)")
```
